Missing the Point$\colon$ Non-Convergence in Iterative Imputation Procedures

### 2020-04-23 2.ResearchReport/1.SimulationStudy/Writeup/NoteConvergence.tex

>>>>

\begin{figure}
%\label{fig:steps2}
\centering
	\large{\begin{tikzpicture}
	\tikzstyle{main}=[circle, minimum size = 25mm, thick, draw =black!80, node distance = 20mm]
\tikzstyle{connect}=[-latex, thick]
\node[main, fill = white!100] (data) [label=below:Incomplete data] {$\{y_{obs}, y_{mis}\}$};
\node[main] (mids) [right=of data] { $\{y_{obs}, y_{imp}\}$};
\node[main] (mids2) [above=10mm of mids] {$\{y_{obs}, y_{imp}\}$ };
\node[main] (mids3) [below=10mm of mids,label=below: Completed data (per $m$)] { $\{y_{obs}, y_{imp}\}$};
\node[main] (mira) [right=of mids] {$\hat{Q}$};
\node[main] (mira2) [above=10mm of mira] {$\hat{Q}$ };
\node[main] (mira3) [below=10mm of mira,label=below: Results (per $m$)] {$\hat{Q}$ };
\node[main, fill = white!100] (mipo) [right=of mira,label=below:Pooled results] {$\bar{Q}$ };
\path (data) edge [connect] (mids)
      (data) edge [connect] (mids2)
      (data) edge [connect] (mids3)
      (mids) edge [connect] (mira)
      (mids2) edge [connect] (mira2)
      (mids3) edge [connect] (mira3)
		  (mira) edge [connect] (mipo)
		  (mira2) edge [connect] (mipo)
		  (mira3) edge [connect] (mipo);
\end{tikzpicture}}
%\caption{Scheme of the main steps in multiple imputation \citep[$m = 3$; adapted from][\S~1.4.1]{buur18}. {\footnotesize Missing data in dataset $Y$ is `imputed' (i.e., filled in) $m$ times. The imputed data is combined with the observed data ($Y_{obs}$) to create $m$ completed datasets. On each completed dataset, the analysis of scientific interest (or `substantive model') is performed. The quantity of scientific interest (e.g., a regression coefficient) is denoted with $Q$. Since $Q$ is estimated on each completed dataset, $m$ separate $\hat{Q}$-values are obtained. These $m$ values are combined into a single pooled estimate $\bar{Q}$. $\{y_{obs}, y_{mis}\}$ }} 
\end{figure}

### 2020-04-23 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd

**Note for peer review: things written in bold are reminders for myself or things that I am unsure off still. Also, I just decided to change things majorly in my results section. So right now, the figures do not correspond to the text. Please only skim that section, because I am definitely going to rewrite it. Thanks already! - Hanne**

@@

Let $y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample. The data value of unit $i$ ($i = 1, \dots, n$) on variable $j$ ($j = 1, \dots, p$) may be either observed or missing. The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$. For each datapoint in $y_{mis}$, we sample $M \times T$ times plausible values, where $M$ is the number of imputations ($m = 1, 2, \dots, M$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$). The collection of samples between the initial value (at $t=1$) and the final imputed value (at $t=T$) will be referred to as an 'imputation chain'. **Add: missingness percentage = proportion of cases with one or more missing values, Q is quantity of scientific interest: "A scientific estimand $Q$ is a quantity of scientific interest that we can calculate if we would observe the entire population" [@buur18, par 2.3.1].  **
<!-- The number of imputations is denoted by $m$ (i.e., the number of times each datapoint in $y_{mis}$ is filled in). The aim of MI is to impute each datapoint in $y_{mis}$ $m$ times. Because of the iterative nature of the MICE algorithm -->

@@

The two challenges of convergence are stationarity and mixing point. Mixing and stationarity can be evaluated visually. Common practice is to evaluate stationarity and mixing in trace plots for scalar summaries of interest. In mice these scalar summaries of interest are chain means and chain variances. As @buur18 describes, users can also specify a model-specific scalar summary (e.g., a regression coefficient). Such a user-specified scalar summary, however, is not universal to all complete data problems. Therefore, as inspired by **someone** (2002), we will use introduce a scalar summary that summarizes the multivariate state space of the algorithm. In this study we use the first eigenvalue of the variance covariance matrix of the completed data set per imputation. The methods under evaluation are $\widehat{R}$ and $AC$. These will be computed for the three scalar summaries of interest: chain means, chain variances, and the first eigenvalue of the variance-covariance matrix across imputations.

The two methods will be evaluated with an estimand per scalar summary of interest. It is assumed that when the estimates are unbiased (and confidence valid), the algorithm is sufficiently converged. The $\widehat{R}$ and $AC$ values with chain means as theta are evaluated against the bias in univariate mean estimates (**just for Y??**). Similarly, the performance measure for Rhat and $AC$ applied to chain variances is the bias in estimated standard deviations. To evaluate the performance of Rhat and $AC$ on the eigenvalues, we will use bias in R squared and in the estimated regression coefficient, the coverage rate of the CI95% of the regression estimate, and the CI length. **remove the term performance measure above and add that performance measures are bias in all estimands, and cov and CI length of regression coeff.**

**Result:** In general there is more bias in the estimate in conditions with a higher proportion of missingness and or a lower number of iterations. As required the values of $\widehat{R}$ are lower for conditions with a higher number off iterations. End somewhat higher in conditions with a higher percentage of missing data. However autocorrelation did not show a decrease with increase number of iterations. After replacing the ACF function with a manual calculation for autocorrelation the autocorrelation values were indeed decreasing with a higher number of iterations. Evaluation with the performance measures shows that are had an autocorrelation are conservative: signs of non-convergence in conditions where estimands are unbiased and or confidence valid. As expected complete convergence (as indicated by autocorrelation equals 0; $\widehat{R}$ equals 1)  is not observed.

@@

Usually, convergence is diagnosed when Rhat < 1.2 or 1.1 or 1.01. And a t-test is performed for $AC$ values == 0. Performance measures are the same as above: unbiased, confidence valid estimates.

**Result:** $\widehat{R}$ is equal to 1.2 is not stringent enough because the performance measures R not unbiased and confidence fellas fell it 4 every condition in which $\widehat{R}$ is smaller than one point 2. At least, not when $\widehat{R}$ is smaller than one point two fur first occurs. The threshold $\widehat{R}$ is equal to one point one is somewhat conservative with respect to the performance measures. It seems OK. $\widehat{R}$ is equal to 1.01 is too stringent compared to the performance measures (**but may be necessary in more complex complete data models???**).

@@

# Results

@@

When applied to the imputation chain means, $\widehat{R}$ indicates that the mice algorithm does not reach a converged state ($\widehat{R}$ = 1) in any of the simulation conditions. Neither is the most recent recommended threshold reached ($\widehat{R}$ < 1.01). The conventional $\widehat{R}$ threshold of 1.2 is reached in simulation conditions $T = 3$ and $T > 6$. (With the default number of iterations (maxit = 5), this dip in Rhat values would be spotted, so it is no problem.) The point at which an extra iteration does not seem to improve the Rhat value is around $T=30$.
<!-- Rhat values are indistinguishable between simulation conditions when $T>30$.  -->
<!-- The $\widehat{R}$ values seemed to flatten off between 20 and 30 iterations.  -->

@@

If I don't include it in my study, also: Significance of $AC$ values; Estimate (PCA or reg. coeff.) versus Rhat and $AC$ on the estimate

>>>>

**Dear Daniela, I have just implemented some major changes in my Methods and Results sections. Therefore, these are kind of a mess. Please do not waste any time on spelling/wording there, I will continue improving it. On the other hand, if you have general advice, please tell me because I'm happy to move thngs around still! Figures in the Results are currently missing, but will help in understanding the paper. Sorry that you won't have that now! Also, things written in bold are reminders for myself or things that I am unsure off still. Thanks already! - Hanne**

@@

Let $y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample. The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, p$) may be either observed or missing. The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$. For each datapoint in $y_{mis}$, we sample $M \times T$ times plausible values, where $M$ is the number of imputations ($m = 1, 2, \dots, M$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$). The collection of samples between the initial value (at $t=1$) and the final imputed value (at $t=T$) will be referred to as an 'imputation chain'. **Add: missingness percentage = proportion of cases with one or more missing values, Q is quantity of scientific interest: "A scientific estimand $Q$ is a quantity of scientific interest that we can calculate if we would observe the entire population" [@buur18, par 2.3.1], and $\theta$s are scalar summaries of interest in the iterative algorithm (e.g., chain means; the average of the imputed values in each imputation chain). **
<!-- The number of imputations is denoted by $m$ (i.e., the number of times each datapoint in $y_{mis}$ is filled in). The aim of MI is to impute each datapoint in $y_{mis}$ $m$ times. Because of the iterative nature of the MICE algorithm -->

@@

The two challenges of convergence are stationarity and mixing. Mixing and stationarity can be inspected visually, by evaluating traceplots of scalar summaries of interest ($\theta$s; e.g., chain means and chain variances). As @buur18 describes, users can also specify a model-specific scalar summary (e.g., a regression coefficient). A user-specified scalar summary, however, is not universal to all complete data problems. Therefore, as inspired by **someone** (2002), we will use introduce a $\theta$ that summarizes the multivariate state space of the algorithm. Namely, the first eigenvalue of the variance-covariance matrix of the $M$ completed datasets.

The methods under evaluation are $\widehat{R}$ and $AC$. These will be computed for the three scalar summaries of interest: chain means (i.e., per imputation, per variable: on $y_{imp, m, j}$), chain variances (i.e., per imputation, per variable: on $y_{imp, m, j}$), and the first eigenvalue of the variance-covariance matrix across imputations (i.e., per imputation: on {$y_{obs}, y_{imp, m}$}) **add user-specified $\theta$: estimated regression coefficient per imputation**.

The performance of two methods ($\widehat{R}$ and $AC$) on each of these scalar summaries of interest will be evaluated with several estimands (quantities of scientific interest; $Q$s). It is assumed that when $\bar{Q}$ (the pooled result across imputations) is an unbiased (and confidence valid) estimate of $Q$, the algorithm is sufficiently converged. For each scalar $\theta$, one estimand $Q$ is defined that may be of interest in empirical research. The $\widehat{R}$ and $AC$ values with chain means as $\theta$s are evaluated against the bias in univariate mean estimates (**all variables or just for Y??**). Similarly, the performance measure for $\widehat{R}$ and $AC$ applied to chain variances is the bias in estimated standard deviations. To evaluate the performance of $\widehat{R}$ and $AC$ on the eigenvalues, we will use bias in the coefficient of determination, $R^2$. Additionally, we will evaluate the estimated regression coefficients, the coverage rate of the CI95% of the regression estimates, and the CI length (**see definitions in old Methods**). **remove the term performance measure here and add an extra alinea to define performance measures as  bias in all estimands, and coverage rate and CI length of regression coefficients.**

**Move this to Results:** As expected, conditions with a higher proportion of missingness and/or a lower number of iterations show more signs of non-convergence, as indicated by more extreme bias in the estimated $Q$s. Roughly speaking this means that MICE is indeed not converged as $t=1$, and converges gradually as $t$ increases. The point at wich an additional iteration does not lead to an improvement of the estimates depends on the difficulty of the missingness problem---between XYZ and XYZ. This means that the algorithm has converged sufficiently (**under the current specifications**).

As required for a method to diagnose non-convergence, the values of $\widehat{R}$ follow a trend similar to the performance measures. $\widehat{R}$ values are generally lower in conditions with a higher number of iterations, and somewhat higher in conditions with a higher percentage of missing data. **Decide where to specify the two calculations of AC:** Initially, autocorrelation did not show a decrease with an increasing number of iterations. But, after replacing the ACF function with a manual calculation for autocorrelation, the autocorrelation values were indeed decreasing with a higher number of iterations.

Evaluation with the performance measures shows that are had an autocorrelation are conservative: they indicate signs of non-convergence in conditions where $\bar{Q}$s are unbiased and or confidence valid estimates of $Q$.

@@

Upon convergence, $\widehat{R}=1$ and $AC=0$, but in MCMC algorithms these are very unlikely (because it's a distribution, not a point to converge to). In practice thresholds are used to diagnose non-convergence. Usually, non-convergence is diagnosed when $\widehat{R}$  > 1.2 or 1.1 or even 1.01. And a t-test is performed to assess whether $AC$ is significantly different from zero. Performance measures are the same as above: unbiased, confidence valid estimates.

**Move this to Results:** As expected, complete convergence ($\widehat{R}=1$ and $AC=0$) is not observed. But what about the thresholds for practice? $\widehat{R}$ <1.2 is not stringent enough, because conditions in which $\widehat{R}$ is smaller than 1.2 do not all yield unbiased estimates of the $Q$s. Moreover, there is a dip in the $\widehat{R}$ values, after which $\widehat{R}$ > 1.2 again. If the algorithms would be terminated at the iteration where $\widehat{R}$ < 1.2 occurs for the first time, the increase in $\widehat{R}$ values might be missed. The threshold $\widehat{R}$ < 1.1 is somewhat conservative with respect to the performance measures. But it seems OK. $\widehat{R}$ < 1.01 is too stringent compared to the performance measures. And it is not obtained for any number of iterations (**but may be necessary in more complex complete data models???**).

@@

# Old Results section

@@

When applied to the imputation chain means, $\widehat{R}$ indicates that the mice algorithm does not reach a converged state ($\widehat{R}$ = 1) in any of the simulation conditions. Neither is the most recent recommended threshold reached ($\widehat{R}$ < 1.01). The conventional $\widehat{R}$ threshold of 1.2 is reached in simulation conditions $T = 3$ and $T > 6$. (With the default number of iterations (maxit = 5), this dip in $\widehat{R}$ values would be spotted, so it is no problem.) The point at which an extra iteration does not seem to improve the $\widehat{R}$ value is around $T=30$.
<!-- $\widehat{R}$ values are indistinguishable between simulation conditions when $T>30$.  -->
<!-- The $\widehat{R}$ values seemed to flatten off between 20 and 30 iterations.  -->

@@

If I don't include it in my study, also: Significance of $AC$ values; Estimate (PCA or reg. coeff.) versus $\widehat{R}$ and $AC$ on the estimate
